Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-03-31 elsevier_MANUSCRIPT.qmd
aes edits

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

@@

-   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
-   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network.
-   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, from global to continental [@beck2013; @santhi2008; @ayers2022; @singh2018]. Such large-scale analyses often utilize generalized datasets and methodologies, resulting in limited applicability to regions with unique hydrogeologic and climatic conditions. Additionally, global and continental-scale studies tend to rely on streamgage networks that disproportionately represent large perennial rivers and watersheds with dense human populations, while underrepresenting arid and semi-arid regions characterized by non-perennial flow regimes and smaller streams [@krabbenhoft-2022]. Thus, their effectiveness in accurately capturing groundwater-surface water interactions, particularly in critically water-stressed dryland regions, remains constrained.

@@

Our data selection criteria ensure a robust analysis but also highlight notable spatial gaps that our machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) covering an area of 295,253 km². For comparison, Indiana (a humid state in the U.S.)maintains 189 active stations within a significantly smaller area of 94,326 km². This results in a streamgage density of approximately 2.004 gauges per 1,000 km² in Indiana, more than three times greater than Arizona's density of 0.623 gauges per 1,000 km². Such disparities in gauge coverage are typical for dryland regions globally [@krabbenhoft-2022], underscoring the necessity and relevance of this type of modeling approach in arid and semi-arid environments.

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

@@

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek.

@@

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|---------------------|----------|----------|----------|----------|----------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|-------------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN). The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly baseflow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.

@@

Analysis of temporal trends indicates a prevailing pattern of declining base flow and BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas. Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity. The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios.

>>>>

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

@@

-   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
-   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse low-flow streamgage network, typical of drylands.
-   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, from global to continental [@beck2013; @santhi2008; @ayers2022; @singh2018]. Such large-scale analyses often utilize generalized datasets and methodologies, resulting in limited applicability to regions with unique hydrogeologic and climatic conditions. Additionally, global and continental-scale studies tend to rely on streamgage networks that disproportionately represent large perennial rivers and regulated watersheds with dense human populations, while underrepresenting arid and semi-arid regions characterized by non-perennial flow regimes and smaller streams [@krabbenhoft-2022]. Thus, their effectiveness in accurately capturing groundwater-surface water interactions, particularly in critically water-stressed dryland regions, remains constrained.

@@

Our data selection criteria ensure a robust analysis but also highlight notable spatial gaps that our machine learning model can address. Arizona has 184 active USGS streamflow stations (as of 2024) covering an area of 295,253 km². For comparison, Indiana (a humid state in the U.S.) maintains 189 active stations within a significantly smaller area of 94,326 km². This results in a streamgage density of approximately 2.004 gauges per 1,000 km² in Indiana, more than three times greater than Arizona's density of 0.623 gauges per 1,000 km². Such disparities in gauge coverage are typical for dryland regions globally [@krabbenhoft-2022], underscoring the necessity and relevance of this type of modeling approach in arid and semi-arid environments.

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

To check for autocorrelation, we applied the Durbin-Watson test [@durbin1950testing], which revealed significant autocorrelation at four streamgages on an annual basis. Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI. This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998]. Trends with a $\rho \le 0.05$ are considered significant.

@@

The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi . The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources. The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023]. High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.

@@

| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
|---------------|------------|------------|------------|------------|------------|
| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |

@@

| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
|---------|---------|---------|---------|---------|---------|---------|---------|
| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |

@@

Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area. @beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN). The U.S. streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006]. Only 10 of these interim gauges overlap geographically with our study area. Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges. Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.

@@

Analysis of temporal trends indicates a prevailing pattern of declining base flow and BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas. Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity. The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios and other stresses to aquifers, such as increased groundwater pumping.
